% Microeconomics*
% Final Exam covering Chs. 1-4, 6-9
% Administered May 2026
%----------------------------------------------
\documentclass[12pt]{exam}

\usepackage{setspace}
\setstretch{1.15}
\nopointsinmargin
\pointformat{}

\usepackage{problemset}

\usepackage[margin=1in]{geometry}

\noprintanswers
%\printanswers

\begin{document}

\FontEasyRead

\Heading
{Microeconomics*}
{Cumulative Test}
{75 Minutes, 100 points plus 5 potential bonus points.
  Annotated (by you) version of my cheat-sheet permitted, and a standard calculator, but
  no other resources. Answer all multiple-choice questions by selecting the best answer.
  Answer all short-response questions clearly, using correct reasoning.
  Show all work for math problems. Do \underline{not} skip any problems - there
is always potential for partial credit.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\ProblemSection{Part I: Multiple Choice (2 points each)}

\begin{questions}

  \begin{mcquestion}[2]
    The ``law of supply'' states that:
    \begin{choices}
      \choice As price increases, quantity supplied decreases
      \choice Supply is independent of price
      \CorrectChoice As price increases, quantity supplied increases
      \choice Supply determines demand
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    A market is purely competitive when:
    \begin{choices}
      \choice All producers and suppliers are price-takers
      \choice There is no distinction between suppliers' product or service
      \choice There is free, uninhibited entry into and exit from in the market
      \CorrectChoice All of the above
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    If income \underline{decreases} and a good is \underline{inferior}, demand for that good will:
    \begin{choices}
      \choice Decrease at first, then increase
      \choice Stay the same
      \CorrectChoice Increase
      \choice Become perfectly elastic
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    Cross-price elasticity of demand measures:
    \begin{choices}
      \choice Responsiveness of change in price of a good or service to a change in the quantity demanded of another good or service
      \choice Responsiveness of quantity demanded of a good or service to a change in its price
      \CorrectChoice Responsiveness of quantity demanded of one good or service to a change in the price of another good or service
      \choice Responsiveness of change in price of a good or service to a change in the quantity demanded
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    If a vendor determines that demand for his product is price \underline{elastic}, he can increase revenue by:
    \begin{choices}
      \choice Increasing price
      \CorrectChoice Decreasing price
      \choice Adding a surcharge
      \choice Leaving price unchanged
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    Average fixed costs:
    \begin{choices}
      \choice Rise with quantity produced
      \choice Remain constant independent of quantity produced
      \choice Are equal to marginal fixed costs
      \CorrectChoice Decline with quantity produced
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    Marginal utility is:
    \begin{choices}
      \choice Total utility divided by quantity
      \CorrectChoice The additional satisfaction gained from consuming one more unit
      \choice A fixed cost per unit
      \choice Equal to average revenue
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    A \underline{shift in demand} for a particular goodcan be caused by:
    \begin{choices}
      \choice An change in income
      \choice A change in the price of the good
      \choice A change in the perceived value of the good
      \CorrectChoice Both A and C
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    Price elasticity of demand is considered \underline{inelastic} when:
    \begin{choices}
      \choice The absolute value of elasticity is greater than 1
      \choice Quantity demanded is highly responsive to price changes
      \CorrectChoice The absolute value of elasticity is less than 1
      \choice Total revenue stays unchanged when price decreases
    \end{choices}
  \end{mcquestion}

  \begin{mcquestion}[2]
    If the price of a \underline{substitute} good increases, demand for the original good will:
    \begin{choices}
      \choice Decrease
      \choice Remain unchanged
      \CorrectChoice Increase
      \choice Become perfectly inelastic
    \end{choices}
  \end{mcquestion}

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \ProblemSection{Part II: Short Answer (10 points each)}

  \begin{shortanswer}[10]{2.5in}
    A sudden 25\% increase in the price of refrigerators causes quantity
    demanded to fall sharply at first, but later recover. Explain what's happening
    in terms of short-run and long-run price elasticity of demand.  Does this imply that refrigerators are a
    durable or a non-durable good?
  \end{shortanswer}
  \begin{solution}
    Short-run demand is more elastic (consumers make-do with their old refrigerators rather than buying
    new ones); long-run demand is more inelastic as consumers' old fridges wear out and they need to buy
    new ones. Refrigerators are durable goods.
  \end{solution}

  \begin{problem}[10]{3in}
    Assuming typical supply and demand curves, fill in each box with the expected effect on
    equilibrium price and quantity. Use these codes:

    \begin{itemize}
      \item Price up, same, down, ambiguous: P-U, P-S, P-D, P-A
      \item Quantity up, same, down, ambiguous:  Q-U, Q-S, Q-D, Q-A
    \end{itemize}

    The easy box has been filled in for you. ``Ambiguous'' here means the outcome could be up, same, or
    down, depending on the relative magnitudes of change and/or the relative elasticities of supply and demand.

    \vspace{0.5cm}

    \begin{center}
      \begin{tikzpicture}[
          every node/.style={draw, align=center, minimum height=1.8cm},
          header/.style={font=\bfseries, fill=gray!15},
          cell/.style={minimum width=3.2cm},
          rowlabel/.style={minimum width=3.2cm, font=\bfseries, fill=gray!15}
        ]
        % \begin{tikzpicture}[
        %     every node/.style={draw, align=center, minimum height=1.8cm},
        %     header/.style={font=\bfseries, fill=gray!15},
        %     cell/.style={minimum width=3.2cm},
        %     rowlabel/.style={minimum width=3.2cm, font=\bfseries, fill=gray!15},
        %     baseline
        %   ]
        %     \begin{tikzpicture}[baseline=(current bounding box.north)]
        % Top row (column headers)
        \node[header, minimum width=3.2cm] at (0,0) {};
        \node[header, cell] at (3.2,0) {Decrease in\\Supply};
        \node[header, cell] at (6.4,0) {No Change\\in Supply};
        \node[header, cell] at (9.6,0) {Increase in\\Supply};

        % Row 1
        \node[rowlabel] at (0,-1.8) {Decrease in\\Demand};
        \node[cell] at (3.2,-1.8) {};
        \node[cell] at (6.4,-1.8) {};
        \node[cell] at (9.6,-1.8) {};

        % Row 2
        \node[rowlabel] at (0,-3.6) {No Change\\in Demand};
        \node[cell] at (3.2,-3.6) {};
        \node[cell] at (6.4,-3.6) {P-S Q-S};
        \node[cell] at (9.6,-3.6) {};

        % Row 3
        \node[rowlabel] at (0,-5.4) {Increase in\\Demand};
        \node[cell] at (3.2,-5.4) {};
        \node[cell] at (6.4,-5.4) {};
        \node[cell] at (9.6,-5.4) {};

      \end{tikzpicture}
    \end{center}

    Sketch curves here to guide your answers (price on vertical axis, quantity on horizontal axis):

  \end{problem}
  \begin{solution}
    Put a solution here.
  \end{solution}

  \begin{shortanswer}[10]{2.5in}
    Explain what an indifference curve is.  Why is it generally downward sloping? Why is
    it generally convex (bowed away from the origin)?
  \end{shortanswer}
  \begin{solution}
    The indifference curve represents all combinations of two goods that yield the same level of
    utility (satisfaction) to a consumer. It is generally downward sloping because if a consumer has
    more of one good, they would need less of the other good to maintain the same level of utility. It
    is convex (bowed away from the origin) due to the principle of diminishing marginal rate of substitution,
    which states that as a consumer substitutes one good for another, they are willing to give up less
    and less of the second good to get additional units of the first good.
  \end{solution}

  \begin{shortanswer}[10]{4in}
    Define the terms \underline{substitute}, \underline{complementary}, and \underline{independent} goods
    in terms of cross-price elasticity of demand. Give an example of each.
  \end{shortanswer}
  \begin{solution}
    Substitute goods have a positive cross-price elasticity of demand, meaning that an increase
    in the price of one good leads to an increase in the demand for the other good. For
    example, if the price of coffee increases, the demand for tea (a substitute) may increase.

    Complementary goods have a negative cross-price elasticity of demand, meaning that an increase
    in the price of one good leads to a decrease in the demand for the other good. For example, if
    the price of printers increases, the demand for printer ink (a complement) may decrease.

    Independent goods have a cross-price elasticity of demand close to zero, meaning that changes
    in the price of one good have little to no effect on the demand for the other good. For example,
    the price of bread may have little effect on the demand for bicycles.
  \end{solution}

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \ProblemSection{Part III: Math Problems (10 points each)}

  \begin{problem}[10]{4in}
    You know a firm's marginal revenue and marginal cost functions in terms of production $q$:
    \[
      MR(q) = 100 - 4q, \quad MC(q) = 20 + q
    \]

    \begin{itemize}
      \item Find its profit-maximizing \( q \)
      \item At that $q$, find price using \( P(q) = 100 - 2q \)
      \item Find total revenue $P \cdot q$
      \item Given a cost function of \( C(q) = 20q + 0.5q^2 \), compute the firm's profit
    \end{itemize}
  \end{problem}
  \begin{solution}
    Set \( MR = MC \):
    \[
      100 - 4q = 20 + q \Rightarrow q = 16
    \]
    \[
      P = 100 - 32 = 68
    \]
    \[
      TR = 68 \cdot 16 = 1088
    \]
    \[
      C = 320 + 128 = 448
    \]
    Profit = 1088 - 448 = 640
  \end{solution}

  \begin{problem}[10]{2in}
    You are given the following (incomplete) CPI-U table (which measures inflation for urban consumers in the U.S.)

    \begin{center}
      \renewcommand{\arraystretch}{1.4}
      \setlength{\tabcolsep}{12pt}

      \begin{tabular}{|>{\centering\arraybackslash}p{3cm}|>{\centering\arraybackslash}p{5cm}|}
        \hline
        \textbf{Calendar Year} & \textbf{CPI-U Annual Average Index} \\
        \textbf{}              & \textbf{(1982--1984 = 100)}         \\
        \hline
        2020                   & 258.811                             \\
        \hline
        2021                   & \entryblank                         \\
        \hline
        2022                   & 292.655                             \\
        \hline
        2023                   & \entryblank                         \\
        \hline
        2024                   & 313.689                             \\
        \hline
        2025                   & 322.180                             \\
        \hline
      \end{tabular}
    \end{center}

    You know that:
    \begin{itemize}
      \item Inflation from 2020 to 2021 was 4.69802\%
      \item \underline{Average annual} inflation (using a discrete compounding basis) from
        2020 to 2023 was 5.59196\%
    \end{itemize}

    Find the two missing CPI values to 3 decimal places.
  \end{problem}
  \begin{solution}
    2021 is 270.970, 2023 is 304.702
  \end{solution}

  \begin{problem}[10]{3in}
    A product price \underline{falls} from 14 to 10 and the quantity demanded \underline{rises}
    from 40 to 60. Compute the observed \underline{arc} price elasticity of demand.  Would you say
    demand for this product is elastic, inelastic, or unit elastic?
  \end{problem}
  \begin{solution}
    $E = \frac{(60-40)/50}{(10-14)/12} = \frac{20/50}{-4/12} = -1.20$
    Since $|E| > 1$, demand is elastic.
  \end{solution}

  \begin{problem}[10]{3in}
    If demand can be described as $P = 50 - 0.5Q$ and total cost $TC = 10 + 20Q$, find the
    profit-maximizing output and price. (Remember that profit is maximized where $MR = MC$).
  \end{problem}
  \begin{solution}
    $TR = (50-0.5Q)Q = 50Q - 0.5Q^2$ $\rightarrow$ $MR = 50 - Q$
    $MC = 20$
    Set $MR = MC: 50 - Q = 20$ $\rightarrow$ $Q = 30$ $\rightarrow$ $P = 35$
  \end{solution}

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \ProblemSection{Bonus Question - 5 points}

  \begin{shortanswer}[5]{3in}
    Identify \underline{one} concept you learned from Microeconomics* that you think helps
    explain real-world pricing, decision-making or policy-setting most clearly to you.  Why is that?
  \end{shortanswer}
  \begin{solution}
    Answers vary!
  \end{solution}

\end{questions}

\end{document}
